prolfqua - R-package for Analysing Proteomics Label Free Quantification Experiments

### 2022-03-18 inst/MyArticle/paper.Rmd
added citations.

Label-Free Quantification (LFQ) Proteomics Experiments enable monitoring relative abundances of thousands of proteins in biological samples. For most studies, parallel group designs are used, e.g., treatment vs. control are used [TODO: cite]. More recently, more complex experimental designs are implemented with increasing numbers of samples, i.e., factorial designs and longitudinal studies (time series), sometimes with repeated measurements on the same subject [TODO: cite]. The obtained data can be modeled using linear fixed-, mixed-, or random-effects models [@Bates2015FittingJSS]. Based on these models, tests can be applied to examine whether a factor or interactions are significant, or differences between factor levels can be estimated, and their significance tested.  Proteins can be ordered or grouped using the estimated fold-change, $t$-statistics, scaled $p$-value, or false discovery rate (FDR) and then subjected to gene set enrichment or over-representation analysis [TODO ...] or validation experiments.

@@

```{r exampleContrasts, fig.cap="(ref:exampleContrasts)", echo=TRUE, out.width = '90%'}
contrasts <- c(
  "dilution_(9/7.5)_1.2" =   "dilution.e - dilution.d",
  "dilution_(7.5/6)_1.25" =   "dilution.d - dilution.c"
)
# fit model
lmmodel <- paste(lfqp$intensity_column()," ~ dilution.")
modelFunction <- strategy_lm( lmmodel, model_name = "lm")
models <- build_model(lfqp, modelFunction)
p0 <- models$anova_histogram()
# compute contrasts from linear model and with imputation
contr <- prolfqua::Contrasts$new(models, contrasts) |> 
  prolfqua::ContrastsModerated$new()
conI <- prolfqua::ContrastsSimpleImpute$new( lfqp, contrasts) |> 
  prolfqua::ContrastsModerated$new()
tmp <- prolfqua::addContrastResults(prefer = contr, add = conI)
# visualize results
pl <- tmp$merged$get_Plotter()
p1 <- pl$histogram()$p.value
p2 <- pl$volcano()$FDR +
   ggplot2::theme(legend.position = "bottom")
gridExtra::grid.arrange(p1, p2, ncol = 2)

>>>>

Label-Free Quantification (LFQ) Proteomics Experiments enable monitoring relative abundances of thousands of proteins in biological samples. For most studies, parallel group designs are used, e.g., treatment vs. control are used [@de2022apoe2,@laubscher2021baf]. More recently, more complex experimental designs are implemented with increasing numbers of samples, i.e., factorial designs and longitudinal studies (time series), sometimes with repeated measurements on the same subject [@tan2022proteomic,@meier2021reduced]. The obtained data can be modeled using linear fixed-, mixed-, or random-effects models [@Bates2015FittingJSS]. Based on these models, tests can be applied to examine whether a factor or interactions are significant, or differences between factor levels can be estimated, and their significance tested.  Proteins can be ordered or grouped using the estimated fold-change, $t$-statistics, scaled $p$-value, or false discovery rate (FDR) and then subjected to gene set enrichment or over-representation analysis [TODO ...] or validation experiments.

@@

```{r exampleContrasts, fig.cap="(ref:exampleContrasts)", echo=TRUE, out.width = '90%'}
contrasts <- c(
  "dilution_(9/7.5)_1.2" =   "dilution.e - dilution.d",
  "dilution_(7.5/6)_1.25" =   "dilution.d - dilution.c"
)
# fit model
lmmodel <- paste(lfqp$intensity_column()," ~ dilution.")
modelFunction <- strategy_lm( lmmodel, model_name = "lm")
models <- build_model(lfqp, modelFunction)
p0 <- models$anova_histogram()
# compute contrasts from linear model and with imputation
contr <- prolfqua::Contrasts$new(models, contrasts) |> 
  prolfqua::ContrastsModerated$new()
conI <- prolfqua::ContrastsSimpleImpute$new( lfqp, contrasts) |> 
  prolfqua::ContrastsModerated$new()
tmp <- prolfqua::addContrastResults(prefer = contr, add = conI)
# visualize results
pl <- tmp$merged$get_Plotter()
p1 <- pl$histogram()$p.value
p2 <- pl$volcano()$FDR +
   ggplot2::theme(legend.position = "bottom")
p1
gridExtra::grid.arrange(p1, p2, ncol = 2)
